Insights of Even-point Multi-loop Unitarity: Exponentiation, Anomalies and Evanescence

### 2023-06-14 eventPointTwoLoop.tex
added a couple references
small changes

\newcommand{\scaleIntBubbleProbeC}[2]{ {
\begin{tikzpicture}[baseline=-2.5]
\begin{feynman}
\vertex (a1) at (-.6,.3) {#1};
\vertex (a2) at (-.6, -.3) {#2};
\vertex (a4) at (.5, .4) {${}^{\alpha_3=D/2-2}$};
\vertex (a6) at (.5, -.5) {${}^{\alpha_4=1}$};
\vertex [dot,scale=2](mid1) at (0,0){};
\vertex [dot,scale=1.5,hgrey0](mid2) at (0,0){};
\vertex [dot,scale=2](mid5) at (1,0){};
\vertex [dot,scale=1.5,hgrey0](mid6) at (1,0){};
\vertex (a3) at (1.6, .3) {#2};
\vertex (a4) at (1.6, -.3) {#2};
\diagram{
(a1) --[ultra thick,](mid1),
(a2) --[ultra thick,](mid1),
(mid1) --[ ultra thick,out=30,in=150,min distance=0.2cm](mid5),
(mid1) --[ ultra thick,out=-30,in=-150,min distance=0.2cm](mid5),
(mid5) --[ultra thick,](a3),
(mid5) --[ultra thick,](a4)
};
\end{feynman}
\end{tikzpicture}
}
}

@@

@@

The remarkable simplicity of one-loop amplitudes is even more dramatic in a fixed spacetime dimension. In $D=4$, the kinematic projection to a basis of scalar integrals saturates at the box integral due to the appearance of Gram determinants. Roughly speaking, $N$-gon integral coefficients of \eqn{TenRed1loop} come dressed with a Gram determinant factor of $\mathcal{G}_N$,
\be
\mathcal{G}_N = \text{det} (k_i \cdot k_j)_{\!N\!\times\! N}
\ee
where $k_i$ are the external momenta flowing into the $N$-gon integral. In fixed spacetime dimension, $D$, the momenta are $D$-component vectors and thus the Lorentz product matrix $(k_i \cdot k_j)_{\!N\!\times\! N}$ must have a null space for $N>D$. Due to this physical constraint, $C_N \neq 0$, only when $N\leq D$ \cite{Bern:1993kr}. We will demonstrate how this can be used in 4D integral construction in the next section.

@@

The existence of closed form expressions, like those in \eqn{eq:integrals}, while common for one-loop integrals, are incredibly rare for higher loop order outside of very specialized kinematic regimes. Luckily, in this work, the integral basis at four-poinnt two-loop order for even-point theories can be reconstructed from the one-loop expressions in \eqn{eq:integrals}. This is a special property of recursively one-loop amplitudes, which we will describe in detail in \sect{sec:recInt}.
\subsection{On-shell Unitarity methods}\label{sec:genU}
Above we have reviewed the known behavior of one-loop integral bases in $D=4$ spacetime dimensions. In this section we will briefly demonstrate why traditional 4D unitarity methods are so powerful -- and why we must go beyond them to accommodate the multi-loop calculations of interest in this work. 
\paragraph{Integrands from 4D Cuts}
Applying one-loop integral reduction in 4D tells us that any one-loop massless amplitude, for which tadpoles are scaleless, can be expressed in terms of box, triangle, and bubble diagrams, with each integral weighted by dimensionally-dependent functions of kinematics:
\be\label{eq:4DUn}
\mathcal{A}^{\text{1-loop}}= \sum_{N=2}^4 C_N(D)I_N^D = \sum_{N=2}^4 C_N(4)I_N^D+ \mathcal{R}
\ee
where the scalar basis-integrals, $I_N^D$, are evaluated in $D=4-2\epsilon$, and $\mathcal{R}$ is a rational term that emerges from series expanding around $\epsilon=(4-D)/2$. To construct integral coefficients, $C_N$, we make use of the Optical Theorem for the $S$-matrix, which states that,
\be
S^\dagger S = \mathbb{I} \quad \Rightarrow \quad 2\,\text{Im}(T) = T^\dagger T
\ee
where $S=\mathbb{I}+iT$. Applying relation to \eqn{eq:4DUn} allows us to identify the imaginary part of $\mathcal{A}^{\text{1-loop}}$ with the imaginary parts of each basis integral weighted by the respective 4D coefficient, $C_N(4)$:
\be\label{eq:UnMeth1loop}
\text{Im}(\mathcal{A}^{\text{1-loop}}) = \sum_{N=2}^4 C_N(4)\, \text{Im}(I_N^D)
\ee
Application of the optical theorem to constrain on-shell amplitudes is known as the unitarity method \cite{UnitarityMethod, BDKUniarityReview}. In order to isolate each $C_N(4)$ appearing in \eqn{eq:4DUn}, we must employ \textit{generalized unitarity}, which places particular internal legs on-shell. That is, for each internal loop propagator, $1/P_i^2$, that we wish to take on-shell, we make the following replacement inside the integral:
\be
\frac{i}{P_i^2 + i\delta} \rightarrow 2\pi \delta(P_i^2)
\ee
This replacement of the internal propagators will select out the requisite imaginary parts needed to determine the integral coefficients. This approach to one-loop amplitude construction was systematized in ref. \cite{Forde:2007mi}, which demonstrated how to directly extract the integrand coefficients directly from 4D cuts:
\begin{align}
C^{(P_iP_jP_k)}_4 &\sim \boxCut
\\
C^{(P_iP_j)}_3 &\sim \triCut - \sum_{k} C^{(P_iP_jP_k)}_4
\\
C^{(P_i)}_2 &\sim \bubCut- \sum_{j} C^{(P_iP_j)} - \sum_{j,k} C^{(P_iP_jP_k)}_4
\end{align}
where exposed legs are summed over all internal states crossing the cut, and the labels $P_i$ are external momentum scales flowing into the specified cuts. The subtractions account for the overlapping information in the respective $N$-gon cuts. By momentum conservation, the momentum flow in the unlabeled vertex is completely determined by the other $N-1$ vertices.

>>>>

\newcommand{\scaleIntBubbleProbeC}[2]{ {
\begin{tikzpicture}[baseline=-2.5]
\begin{feynman}
\vertex (a1) at (-.6,.3) {#1};
\vertex (a2) at (-.6, -.3) {#2};
\vertex (a4) at (.5, .4) {${}^{\alpha_3=D/2-2}$};
\vertex (a6) at (.5, -.5) {${}^{\alpha_4=1}$};
\vertex [dot,scale=2](mid1) at (0,0){};
\vertex [dot,scale=1.5,hgrey0](mid2) at (0,0){};
\vertex [dot,scale=2](mid5) at (1,0){};
\vertex [dot,scale=1.5,hgrey0](mid6) at (1,0){};
\vertex (a3) at (1.6, .3) {#2};
\vertex (a4) at (1.6, -.3) {#2};
\diagram{
(a1) --[ultra thick,](mid1),
(a2) --[ultra thick,](mid1),
(mid1) --[ ultra thick,out=30,in=150,min distance=0.3cm](mid5),
(mid1) --[ ultra thick,out=-30,in=-150,min distance=0.3cm](mid5),
(mid5) --[ultra thick,](a3),
(mid5) --[ultra thick,](a4)
};
\end{feynman}
\end{tikzpicture}
}
}

@@

\newcommand{\pentInt}{ {
\begin{tikzpicture}[baseline=-2]
\begin{feynman}
\vertex (a1) [dot,scale=1.6] at (0,.65){};
\vertex (a2) [dot,scale=1.6]at (.55,.2){};
\vertex (a3) [dot,scale=1.6]at (.3,-.4){};
\vertex (a4) [dot,scale=1.6]at (-.3,-.4){};
\vertex (a5) [dot,scale=1.6]at (-.55,.2){};
\vertex (a1) [dot,hgrey0,scale=1.3] at (0,.65){};
\vertex (a2) [dot,hgrey0,scale=1.3]at (.55,.2){};
\vertex (a3) [dot,hgrey0,scale=1.3]at  (.3,-.4){};
\vertex (a4) [dot,hgrey0,scale=1.3]at (-.3,-.4){};
\vertex (a5) [dot,hgrey0,scale=1.3]at (-.55,.2){};
\vertex (b4) at (.5,-.6);
\vertex (b3) at (-.5,-.6);
\vertex (b2) at (-.8,.3);
\vertex (b1) at (0,.9) ;
\vertex (b5) at (.8,.3) ;
\diagram{
(a2) --[ ultra thick,](a1),
(a3) --[ ultra thick,](a2),
(a4) --[ ultra thick,](a3),
(a5) --[ ultra thick,](a4),
(a5) --[ ultra thick,](a1),
(b1) --[ ultra thick,](a1),
(b2) --[ ultra thick,](a5),
(b3) --[ ultra thick,](a4),
(b4) --[ ultra thick,](a3),
(b5) --[ ultra thick,](a2),
};
\end{feynman}
\end{tikzpicture}
}
}

\newcommand{\boxPinch}{ {
\begin{tikzpicture}[baseline=-1]
\begin{feynman}
\vertex (a1) [dot,scale=1.6] at (.4,.4){};
\vertex (a2) [dot,scale=1.6]at (-.4,.4){};
\vertex (a3) [dot,scale=1.6]at (-.4,-.4){};
\vertex (a4) [dot,scale=1.6]at (.4,-.4){};
\vertex (a1) [dot,hgrey0,scale=1.3] at (.4,.4){};
\vertex (a2) [dot,hgrey0,scale=1.3]at (-.4,.4){};
\vertex (a3) [dot,hgrey0,scale=1.3]at (-.4,-.4){};
\vertex (a4) [dot,hgrey0,scale=1.3]at (.4,-.4){};
\vertex (b4) at (.6,-.6);
\vertex (b3) at (-.6,-.6);
\vertex (b2) at (-.6,.6);
\vertex (b1a) at (.6,.75) ;
\vertex (b1b) at (.75,.6) ;
\vertex (c1) at (.7,.9) {${}^{i}$};
\vertex (c2) at (1.1,.6) {${}^{i+1}$};
\diagram{
(a2) --[ ultra thick,](a1),
(a3) --[ ultra thick,](a2),
(a4) --[ ultra thick,](a3),
(a1) --[ ultra thick,](a4),
(a2) --[ ultra thick,](b2),
(a3) --[ ultra thick,](b3),
(a4) --[ ultra thick,](b4),
(a1) --[ ultra thick,](b1a),
(a1) --[ ultra thick,](b1b),
};
\end{feynman}
\end{tikzpicture}
}
}

@@

The remarkable simplicity of one-loop amplitudes is even more dramatic in a fixed spacetime dimension. In $D=4$, the kinematic projection to a basis of scalar integrals saturates at the box integral due to the appearance of Gram determinants. Roughly speaking, $(N+1)$-gon integral coefficients of \eqn{TenRed1loop} come dressed with a Gram determinant factor of $\mathcal{G}_N$,
\be
\mathcal{G}_N = \text{det} (k_i \cdot k_j)_{\!N\!\times\! N}
\ee
where $k_i$ are the $N$ linearly independent external momenta flowing into the $(N+1)$-gon integral. In fixed spacetime dimension, $D$, the momenta are $D$-component vectors and thus the Lorentz product matrix $(k_i \cdot k_j)_{\!N\!\times\! N}$ must have a null space for $N>D$. Due to this physical constraint, $C_M \neq 0$, only when $M\leq D+1$. Furthermore, in $D=4-2\epsilon$ dimensions, any scalar pentagon integral can be rewritten up to $\mathcal{O}(\epsilon)$ in terms of five pinched scalar box integrals \cite{Bern:1993kr}. Due to this property of 4D pentagon integrals, all the information of one-loop 4D amplitudes can be completely determined by evaluating up to box integrals, regardless of the multiplicity. We will demonstrate in the next section how this can be used in 4D integral construction.

@@

The existence of closed form expressions, like those in \eqn{eq:integrals}, while common for one-loop integrals, are incredibly rare for higher loop order outside of very specialized kinematic regimes. Luckily, in this work, the integral basis at four-poinnt two-loop order for even-point theories can be reconstructed from the one-loop expressions in \eqn{eq:integrals}. This is a special property of recursively one-loop amplitudes, which we will describe in detail in \sect{sec:recInt}.
\subsection{On-shell Unitarity methods}\label{sec:genU}
Above we have reviewed the known behavior of one-loop integral bases in $D=4$ spacetime dimensions. In this section we will briefly demonstrate why traditional 4D unitarity methods are so powerful -- and why we must go beyond them to accommodate the multi-loop calculations of interest in this work. 
\paragraph{Integrands from 4D Cuts}
Applying one-loop integral reduction in 4D tells us that any one-loop massless amplitude, for which tadpoles are scaleless, can be expressed in terms of box, triangle, and bubble diagrams, with each integral weighted by dimensionally-dependent functions of kinematics:
\be\label{eq:4DUn}
\mathcal{A}^{\text{1-loop}}= \sum_{N=2}^4 C_N(D)I_N^D = \sum_{N=2}^4 C_N(4)I_N^D+ \mathcal{R}
\ee
where the scalar basis-integrals, $I_N^D$, are evaluated in $D=4-2\epsilon$, and $\mathcal{R}$ is a rational term that emerges from series expanding around $\epsilon=(4-D)/2$. To construct integral coefficients, $C_N$, we make use of the Optical Theorem for the $S$-matrix, which states that,
\be
S^\dagger S = \mathbb{I} \quad \Rightarrow \quad 2\,\text{Im}(T) = T^\dagger T
\ee
where $S=\mathbb{I}+iT$. Applying relation to \eqn{eq:4DUn} allows us to identify the imaginary part of $\mathcal{A}^{\text{1-loop}}$ with the imaginary parts of each basis integral weighted by the respective 4D coefficient, $C_N(4)$:
\be\label{eq:UnMeth1loop}
\text{Im}(\mathcal{A}^{\text{1-loop}}) = \sum_{N=2}^4 C_N(4)\, \text{Im}(I_N^D)
\ee
Application of the optical theorem to constrain on-shell amplitudes is known as the unitarity method \cite{UnitarityMethod, BDKUniarityReview}. In order to isolate each $C_N(4)$ appearing in \eqn{eq:4DUn}, we must employ \textit{generalized unitarity}, which places particular internal legs on-shell. That is, for each internal loop propagator, $1/P_i^2$, that we wish to take on-shell, we make the following replacement inside the integral:
\be
\frac{i}{P_i^2 + i\delta} \rightarrow 2\pi \delta(P_i^2)
\ee
This replacement of the internal propagators will select out the requisite imaginary parts needed to determine the integral coefficients. This approach to one-loop amplitude construction was systematized in ref.~\cite{Forde:2007mi}, which demonstrated how to directly extract the integrand coefficients directly from 4D cuts:
\begin{align}
C^{(P_iP_jP_k)}_4 &\sim \boxCut
\\
C^{(P_iP_j)}_3 &\sim \triCut - \sum_{k} C^{(P_iP_jP_k)}_4
\\
C^{(P_i)}_2 &\sim \bubCut- \sum_{j} C^{(P_iP_j)} - \sum_{j,k} C^{(P_iP_jP_k)}_4
\end{align}
where exposed legs are summed over all internal states crossing the cut, and the labels $P_i$ are external momentum scales flowing into the specified cuts. The subtractions account for the overlapping information in the respective $N$-gon cuts. By momentum conservation, the momentum flow in the unlabeled vertex is completely determined by the other $N-1$ vertices.
